OECD
========================================================
```{r load, cache=TRUE}
library(ggplot2)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)


# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)
```

Average test scores: each student has 5 scores for each area.  
```{r avg, cache=TRUE}
avg_test = function(subj) {
  apply(student2012[, paste0("PV", 1:5, subj)], 1, mean)
}
subj = c("MATH", "MACC", "MACQ", "MACS", "MACU", "MAPE", "MAPF", "MAPI", "READ", "SCIE")

#rank avg math scores by country
student2012$math = avg_test("MATH")

math_scores = student2012 %>% 
    select(CNT, math) %>% 
    group_by(CNT) %>% 
    summarise(m_math = mean(math)) %>% 
    arrange(m_math)

#rank avg reading scores by country
student2012$read = avg_test("READ")

read_scores = student2012 %>% 
    select(CNT, read) %>% 
    group_by(CNT) %>% 
    summarise(m_read = mean(read)) %>% 
    arrange(m_read)

#rank avg science scores by country
student2012$sci = avg_test("SCIE")

sci_scores = student2012 %>% 
    select(CNT, sci) %>% 
    group_by(CNT) %>% 
    summarise(m_sci = mean(sci)) %>% 
    arrange(m_sci)

print(math_scores)
print(read_scores)
print(sci_scores)
```

I've heard that Chinese students on average spend the most time studying and Finnish students spend the least amount of time.  We can see that Finnish students study very little outside of school.
```{r hours, cache=TRUE}
#removing the stupid attributes from atomic vectors
student2012$MMINS = as.numeric(student2012$MMINS)
student2012$LMINS = as.numeric(student2012$LMINS)
student2012$SMINS = as.numeric(student2012$SMINS)
student2012$ST57Q01 = as.numeric(student2012$ST57Q01)
student2012$ST57Q02 = as.numeric(student2012$ST57Q02)
student2012$ST57Q03 = as.numeric(student2012$ST57Q03)
student2012$ST57Q04 = as.numeric(student2012$ST57Q04)
student2012$ST57Q05 = as.numeric(student2012$ST57Q05)

#all time related vars
dict_student2012[grep("time", dict_student2012[,2], ignore.case=TRUE), ]

OUTHOURS = student2012 %>%
    select(CNT, LMINS, MMINS, OUTHOURS, SMINS, ST57Q01, ST57Q02, ST57Q03, ST57Q04, ST57Q05) %>%
    group_by(CNT) %>%
    summarise(m_OUTHOURS = mean(OUTHOURS, na.rm=T)) %>%
    arrange(m_OUTHOURS)
print(OUTHOURS)
```

We can calculate score per hour to normalize test scores against how much students actually study.
```{r norm, cache=TRUE}
OUTHOURS = OUTHOURS %>% 
  group_by(CNT) %>% 
  left_join(math_scores) %>% 
  left_join(read_scores) %>% 
  left_join(sci_scores) %>%
  mutate(math_hour = m_math / m_OUTHOURS) %>%
  mutate(read_hour = m_read / m_OUTHOURS) %>%
  mutate(sci_hour = m_sci / m_OUTHOURS)

print(OUTHOURS[, c(1,6:8)])
```

Let's see if parent income has any effect on student performance.  No surprise here.
```{r parents, cache=TRUE}
library(ggvis)
students = student2012 %>% select(CNT, STIDSTD, math, read, sci)
parents = parent2012 %>% select(CNT, STIDSTD, PA07Q01,
                                PA03Q01, PA03Q02, PA03Q03, PA03Q04,
                                PA05Q01, PA05Q02, PA05Q03, PA05Q04)
student_parent = students %>% left_join(parents, by = c("STIDSTD", "CNT"))
student_parent = student_parent[!is.na(student_parent$PA07Q01), ]
qplot(x = PA07Q01, y = math, geom="boxplot", data = student_parent)
```

```{r time, cache=TRUE}
time_vars = c(paste0("ST57Q0", 1:6), "LMINS", "MMINS", "SMINS", "OUTHOURS", paste0("ST69Q0", 1:3), paste0("ST70Q0", 1:3), "ST71Q01")

#  ST57Q01                                       Out-of-School Study Time - Homework
#  ST57Q02                                Out-of-School Study Time - Guided Homework
#  ST57Q03                                 Out-of-School Study Time - Personal Tutor
#  ST57Q04                             Out-of-School Study Time - Commercial Company
#  ST57Q05                                    Out-of-School Study Time - With Parent
#  ST57Q06                                       Out-of-School Study Time - Computer
#172 ST69Q01	Min in <class period> - <test lang>
#173	ST69Q02	Min in <class period> - <Maths>
#174	ST69Q03	Min in <class period> - <Science>
#175	ST70Q01	No of <class period> p/wk - <test lang>
#176	ST70Q02	No of <class period> p/wk - <Maths>
#177	ST70Q03	No of <class period> p/wk - <Science>
#178	ST71Q01	No of ALL <class period> a week

```