Single parents
==============

```{r load, echo=FALSE, results='hide'}
library(ggplot2)
library(plyr)
library(dplyr)
setwd("..")
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name) {

    template <- c("2012.rda", "2012dict.rda")

    file_name %>% vapply(paste0, template, template) %>% file.path(".", "data", 
        .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)


# function to convert to data-frames
fn_make_df <- function(named_vector) {
    data.frame(variable = attr(named_vector, "names"), description = named_vector, 
        row.names = NULL)
}

dict_item2012 <- fn_make_df(item2012dict)
dict_parent2012 <- fn_make_df(parent2012dict)
dict_school2012 <- fn_make_df(school2012dict)
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict)
dict_student2012 <- fn_make_df(student2012dict)

rm(fn_make_df, fn_build, sets, item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)

```


Grandparents at home doesn't have much of an effect on scores.  If we look at the average test scores (averaging math, science and reading), there is an obvious difference between have both parents versus have neither parent.  However for single parents, having only a father will generally mean a much lower score.
```{r singleparents}
athome = student2012 %>% select(CNT, SCHOOLID, STIDSTD, OECD, PV1MATH, PV1SCIE, PV1READ,
                                ST11Q01, ST11Q02, ST11Q03,
                                ST11Q04, ST11Q05, ST11Q06)
athome$PV1MATH = as.numeric(athome$PV1MATH)
athome$PV1SCIE = as.numeric(athome$PV1SCIE)
athome$PV1READ = as.numeric(athome$PV1READ)
athome$CNT = as.character(athome$CNT)
athome$CNT[athome$CNT=="Serbia"] <- "Republic of Serbia"
athome$CNT[athome$CNT=="Korea"] <- "South Korea"
athome$CNT[athome$CNT=="Chinese Taipei"] <- "Taiwan"
athome$CNT[athome$CNT=="Slovak Republic"] <- "Slovakia"
athome$CNT[athome$CNT=="Russian Federation"] <- "Russia"
athome$CNT[athome$CNT=="Perm(Russian Federation)"] <- "Russia"
athome$CNT[athome$CNT=="Hong Kong-China"] <- "Hong Kong S.A.R."
athome$CNT[athome$CNT=="China-Shanghai"] <- "China"
athome$CNT[athome$CNT=="China-Macau"] <- "China"
athome$CNT[athome$CNT=="Connecticut (USA)"] <- "United States of America"
athome$CNT[athome$CNT=="Florida (USA)"] <- "United States of America"
athome$CNT[athome$CNT=="Massachusetts (USA)"] <- "United States of America"


#########################################
#overall
overall = athome %>% group_by(CNT) %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_overall = (mmath + mread + msci) / 3) %>% arrange(desc(overall_overall))

#both parents
both = athome %>% group_by(CNT) %>% filter(ST11Q01 == "Yes" & ST11Q02 == "Yes") %>%
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_both = (mmath + mread + msci) / 3) %>% arrange(desc(overall_both))

#single mothers
mother = athome %>% group_by(CNT) %>% filter(ST11Q01 == "Yes" & ST11Q02 == "No") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_mother = (mmath + mread + msci) / 3) %>% arrange(desc(overall_mother))

#single fathers
father = athome %>% group_by(CNT) %>% filter(ST11Q01 == "No" & ST11Q02 == "Yes") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_father = (mmath + mread + msci) / 3) %>% arrange(desc(overall_father))

#neither mother nor father
neither = athome %>% group_by(CNT) %>% filter(ST11Q01 == "No" & ST11Q02 == "No") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_neither = (mmath + mread + msci) / 3) %>% arrange(desc(overall_neither))

#grandparents
grand = athome %>% group_by(CNT) %>% filter(ST11Q05 == "Yes") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_grand = (mmath + mread + msci) / 3) %>% arrange(desc(overall_grand))

#no grandparents
nogrand= athome %>% group_by(CNT) %>% filter(ST11Q05 == "No") %>% 
  summarise(mmath = mean(PV1MATH), msci = mean(PV1SCIE), mread = mean(PV1READ), n = length(CNT)) %>%
  mutate(overall_grand = (mmath + mread + msci) / 3) %>% arrange(desc(overall_grand))

#########################################
athome = athome %>% filter(!is.na(athome$ST11Q01) & !is.na(athome$ST11Q02))
athome$parents = NA
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "Yes", ]$parents = "both"
athome[athome$ST11Q01 == "Yes" & athome$ST11Q02 == "No", ]$parents = "mother"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "Yes", ]$parents = "father"
athome[athome$ST11Q01 == "No" & athome$ST11Q02 == "No", ]$parents = "neither"
#qplot(x = parents, y = PV1MATH, geom = "boxplot", facets = ~CNT, data = athome) + coord_flip()



single_parents = merge(mother[,c(1,6)], father[,c(1,6)], by.x = c("CNT"), by.y = "CNT")
single_parents$diff = single_parents$overall_mother - single_parents$overall_father
single_parents = single_parents %>% arrange(desc(diff))
single_parents$CNT = factor(single_parents$CNT,
                            levels = single_parents$CNT[order(single_parents$diff)])
single_parents %>% left_join(overall %>% select(CNT, overall_overall))
df = merge(single_parents, overall %>% select(CNT, overall_overall), by.x = 'CNT', by.y = 'CNT')
df = merge(df, athome %>% group_by(CNT) %>% summarise(oecd = OECD[1]))
#single mother scores - single father scores
qplot(CNT, diff, data = df, color = oecd) + coord_flip()
```

Comparing having both parents to just having a mother, it appears there are many countries where there isn't a huge difference in test scores.
```{r single-both, eval=FALSE}
#both parents to single mothers only
df2 = merge(mother[,c(1,6)], both[, c(1,6)], by.x = c("CNT"), by.y = c("CNT"))
df2$diff = df2$overall_both - df2$overall_mother
df2 %>% arrange(desc(diff))
```
