# First pass through the data

This will be a stream-of-concoiusness walk through the data, a first pass to see what I can see.

## Libraries

At some point, we should make sure we agree on which versions of which packages we propose to use. As a starting point, I can show what I am using:

```{r load_libraries, message=FALSE}
library(stringr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggvis)
```

Here's my versions:

```{r session_info}
sessionInfo()
```

I suspect that for publication, `ggplot2` will be more capable than `ggvis`.

# Load the data

```{r load_data}
sets <- c("item", "parent", "school", "scoredItem", "student")

# function to build the file names
fn_build <- function(file_name){
 
  template <- c("2012.rda", "2012dict.rda")
 
  file_name %>% 
    vapply(str_join, template, template) %>% 
    file.path("..", "data", .)
}

# load the data
sets %>% fn_build %>% lapply(load, .GlobalEnv)

# clean
rm(fn_build, sets)
```

The dictionaries are stored as named vectors - I would prefer data frames so that I can use the RStudio GUI to examine them.

```{r modify_dictionaries}
# function to convert to data-frames
fn_make_df <- function(named_vector){
  data.frame(
    variable = attr(named_vector, "names"),
    description = named_vector,
    row.names = NULL
  )
}

# there's a clever way to do this, but beyond me for naw
dict_item2012 <- fn_make_df(item2012dict) 
dict_parent2012 <- fn_make_df(parent2012dict) 
dict_school2012 <- fn_make_df(school2012dict) 
dict_scoredItem2012 <- fn_make_df(scoredItem2012dict) 
dict_student2012 <- fn_make_df(student2012dict) 

# clean
rm(fn_make_df)
rm(item2012dict, parent2012dict, school2012dict, scoredItem2012dict, student2012dict)
```

Looking at the number of variables in each of the data frames, I wish I knew how to use ggobi...

One of the tracks to investigate is inequality. I see there is a package, `ineq` on CRAN, that claims to have a Gini-coefficient function, so that seems like a good place to start.

```{r ineq}
library(ineq)
```

Let's get an idea, for each country and subject, an estimate of the inequality of outcomes.

NOTE: `dplyr` seems to choke on grouping of data.frames with NA, hence the `filter()`.

Also: is there a problem the help text for `dplyr::grouped_df`? The description of `drop` seems backwards to me.

```{r twiddle}
df_twiddle <- student2012 %>%
  select(country = CNT, age = ST06Q01,
         math = PV1MATH, reading = PV1READ, science = PV1SCIE) %>%
  gather(key = subject, value = score, -country, -age) %>%
  do(filter(., complete.cases(.))) %>%
  dplyr::group_by(country, subject) %>%
  summarize(
    n = n(), 
    gini_coeff = ineq(score, type="Gini"), 
    mean = mean(score), 
    median = median(score)) %>%
  arrange(as.character(country))
```

"Alabama first" works to look for a specific country, but lets arrange by something more meaningful. 

Let's get a mean of coefficents over all subjects, to help us order things.

```{r by_country}
df_twiddle_mean <- df_twiddle %>%
  group_by(country) %>%
  summarize(
    gini_coeff = mean(gini_coeff),
    mean = mean(mean),
    median = mean(median)
  )
```


```{r arrange}
my_levels <- 
  df_twiddle_mean$country[rev(order(df_twiddle_mean$gini_coeff))] %>%
  as.character

df_twiddle$country <- factor(
  df_twiddle$country,
  levels = my_levels
)

head(my_levels)

tail(my_levels)
```

I know I should figure out the `ggmap` package (I'll look at Di's code), but in the meantinme:

```{r vis_ineq_median}
ggplot(
  df_twiddle %>% filter(subject != "mean_subject"), 
  aes(x = gini_coeff, y = median, color = subject, size = n)
) + 
geom_point(alpha = 0.75) + 
scale_size_area()
```

This is print, so using ggvis and toolips is out of the question...

There appear two clusters - it will be interesting to see if there is a thread to bind them.

```{r vis_countries}
ggplot(
  df_twiddle,
  aes(x = gini_coeff, y = country, color = subject, size = n)
) + 
geom_point() + 
scale_size_area()

ggplot(
  df_twiddle,
  aes(x = median, y = country, color = subject, size = n)
) + 
geom_point() + 
scale_size_area()
```

Hopefully enough to get started.




